---
title: "'Dispersion' in corpus linguistics and statistics"
description: "This blog post clarifies the different ways in which the term *dispersion* is used in corpus linguistics and statistics."
date: 2023-11-16
categories: [corpus linguistics, statistics, dispersion, terminology]
citation: 
  url: https://lsoenning.github.io/posts/2023-01-18_dispersion_terminology/
draft: false
---

```{r}
#| code-fold: true
#| code-summary: "R setup"
#| message: false
#| warning: false

library(lattice)

source("C:/Users/ba4rh5/Work Folders/My Files/R projects/my_utils_website.R")
```

In corpus linguistics, the term *dispersion* is used to describe the distribution of an item or structure in a corpus [see @Gries2008; @Gries2020]. For most dispersion measures, a corpus must first be divided into units (or parts). These units commonly reflect the design of the corpus -- they can be text files, for instance, or text categories. A dispersion index then describes the distribution of an item across these units. There are two general classes of measures:

-   those measuring the *pervasiveness* of an item, which is reflected in the number of units that contain the item (Range and Text Dispersion, its proportional analog)
-   the much larger class of *evenness* measures, which express how evenly an item is distributed across the units (e.g. *D*, *D~2~*, *S*, *D~P~*, *D~A~*, *D~KL~*).

Most dispersion measures range between 0 and 1, where 1 indicates a perfectly even distribution, or the maximal degree of pervasiveness (i.e. the item occurs in every unit).

From a statistical viewpoint, the input for the calculation of evenness measures would be considered a count variable, since it records the number of events (occurrences of the item) that are observed during a certain period of observation. In corpus linguistics, the "period of observation" is "text time", expressed as a word count.

There is an extensive literature on the use of regression models for count variables [e.g. @Long1997; @Cameron_Trivedi2013; @Hilbe2014], and such models have seen some successful applications to word frequency data [e.g. @Mosteller_Wallace1984; @Church_Gale1995]; @Winter_Buerkner2021 provide an accessible introduction for linguists. In this literature, the term "dispersion" is also used, though with a different (apparently opposite) meaning.

Let us first consider the corpus-linguistic (and lexicographic) sense, which can be best described visually, using a so-called "dispersion plot". @fig-1 shows a dispersion plot for two corpora, A and B. The framed rectangles represent the sequence of words forming the corpus, and the spikes inside of these locate the occurrences of a specific item in the corpus. In corpus A, the item is spread out quite evenly. In corpus B, instances are more densely clustered, and there are large stretches where the item does not occur. In the corpus-linguistic sense, then, the dispersion of the item is greater in corpus A. The dispersion score for the item would be greater in Corpus A (i.e. closer to 1).

```{r fig.height=1.75, fig.width=5}
#| fig-cap: "Dispersion in the corpus-linguistic sense: Distribution of word tokens in the corpus."
#| label: fig-1
#| fig-height: 1.75
#| fig-width: 5
#| code-fold: true
#| code-summary: "R code: Figure 1"

set.seed(2000)

n_tokens_A <- c(3,5,4,4,3,4,4,5)
n_tokens_B <- c(5,0,1,9,0,1,0,3)

n_texts <- length(n_tokens_A)

A_loc <- rep(1:n_texts, n_tokens_A)+runif(sum(n_tokens_A))
B_loc <- rep((1:n_texts)[n_tokens_B!=0], n_tokens_B[n_tokens_B!=0])+runif(sum(n_tokens_B))

xyplot(
  1~1, type="n", xlim=c(-1,14), ylim=c(2.8,6),
  par.settings=my_settings, scales=list(draw=F), xlab="", ylab="",
  panel=function(...){
    panel.rect(xleft=c(1,3,5,7), xright=c(2,4,6,8), ytop=5.1, ybottom=4.75, 
               border=FALSE, col="grey90", lwd=1)
    panel.rect(xleft=1, xright=n_texts+1, ytop=5.1, ybottom=4.75, 
               border="grey50", lwd=1)
    

        
    panel.segments(x0=A_loc, x1=A_loc, y0=4.8, y1=5.05, lwd=.75)
    panel.text(x=(1:n_texts)+.5, y=4.55, label=n_tokens_A, 
               col="grey50", cex=.9)
    
    
    panel.rect(xleft=c(1,3,5,7), xright=c(2,4,6,8), ytop=4.1, ybottom=3.75, 
               border=FALSE, col="grey90", lwd=1)
    
    panel.rect(xleft=1, xright=n_texts+1, ytop=4.1, ybottom=3.75, 
               border="grey60", lwd=1)
    
    panel.segments(x0=B_loc, x1=B_loc, y0=3.8, y1=4.05, lwd=.75)
    
    panel.text(x=(1:n_texts)+.5, y=3.55, label=n_tokens_B, 
               col="grey60", cex=.9)
    
    panel.text(x=.4, y=c(4,5)-.1, label=c("Corpus B", "Corpus A"), 
               adj=1, cex=1)
    
    panel.text(x=9.7, y=c(3.8, 4.8), adj=0, 
               label=c("Lower dispersion\n", "Higher dispersion\n"), 
               col=1, lineheight=.85, cex=1)
    panel.text(x=9.7, y=c(3.8, 4.8), adj=0, 
               label=c("\n(more concentrated)", "\n(more spread out)"), 
               col="grey40", lineheight=.85, cex=.9)
    
    panel.segments(x0=1, x1=2, y0=5.4, y1=5.4, lwd=.5, col="grey50")
    panel.segments(x0=1:2, x1=1:2, y0=5.4, y1=5.3, lwd=.5, col="grey50")
    
    
    panel.text(x=1.5, y=5.75, label="Text 1", col="grey40", cex=.8)
    panel.text(x=7, y=2.75, label="Occurrences\nof item in text", col="grey40", 
               cex=.9, lineheight=.85)
    panel.segments(x0=5.9, x1=5.6, y0=3, y1=3.3, col="grey40", lwd=.5)
    })

```

Note how each corpus is divided into 8 texts, which are shown in @fig-1 using greyshading. The numbers below the dispersion plot for each corpus report the number of occurrences of the item in each text. For corpus A, they range between 3 and 5; for corpus B, between 0 and 9.

@fig-2 shows a different representation of these data. Instead of looking at the corpus as a string of words, we consider the text-specific frequencies (sometimes called sub-frequencies) of the item. These indicate how often the item occurs in each document. @fig-2 shows these text-level token counts: Each text is represented by a dot, which marks how often the item appears in the text. In our hypothetical corpora, each text has the same length, which is why we can compare absolute counts. If texts differ in length, we would instead use normalized frequencies, i.e. occurrence rates such as "3.1 per thousand words".

```{r fig.height=1.5, fig.width=3.5}
#| fig-cap: "Dispersion in the statistical sense: Distribution of text-level ocurrence rates."
#| label: fig-2
#| fig-height: 1.5
#| fig-width: 3.5
#| code-fold: true
#| code-summary: "R code: Figure 2"
#| classes: preview-image

xyplot(
  1~1, type="n", xlim=c(-1,7.5), ylim=c(-.35,2.3),
  par.settings=my_settings, scales=list(draw=F), xlab="", ylab="",
  panel=function(...){
    panel.dotdiagram(1+(n_tokens_A/5), y_anchor=1, scale_y=.125, set_cex=1.3)
    panel.dotdiagram(1+(n_tokens_B/5), y_anchor=0, scale_y=.125, set_cex=1.3)
    panel.segments(x0=1, x1=3.2, y0=1, y1=1)
    panel.segments(x0=1, x1=3.2, y0=0, y1=0)
    panel.segments(x0=1+c(0,5,10)/5, x1=1+c(0,5,10)/5, y0=1, y1=.95)
    panel.segments(x0=1+c(0,5,10)/5, x1=1+c(0,5,10)/5, y0=0, y1=-.05)
    panel.text(x=1+c(0,5,10)/5, y=-.2, label=c(0,5,10), col="grey40", cex=.8)
    
    panel.text(x=.6, y=c(0,1)+.1, label=c("Corpus B", "Corpus A"), 
               adj=1, cex=1)
    
    panel.text(x=4, y=c(0.1 ,1.1), adj=0, 
               label=c("Higher dispersion\n", "Lower dispersion\n"), 
               col=1, lineheight=.85, cex=1)
    panel.text(x=4, y=c(0.1 ,1.1), adj=0, 
               label=c("\n(more spread out)", "\n(more concentrated)"), 
               col="grey40", lineheight=.85, cex=.9)
    
    panel.text(x=2, y=-.5, label="Occurrences of item", cex=1, lineheight=.9)
    panel.text(x=3, y=2.2, label="Each dot\nrepresents a text", cex=.9, 
               lineheight=.9, col="grey40")
    })
```

If we compare the distribution of text-level occurrence rates in the two corpora, we note that while the texts in corpus A form a dense pile, the occurrence rates in corpus B are more widely spread out. At this level of description, then, it is the data from corpus B that show greater "dispersion". In the statistical literature on count regression, the term dispersion is used in this sense, i.e. to refer to the variability of unit-specific (i.e. text-level) occurrence rates [e.g. @Long1997, p. 221; @Gelman_etal2021, p. 264--268]. An awareness of the different meanings of "dispersion" will prove helpful for corpus linguists (and lexicographers) when engaging with the statistical literature on count data modeling.

::: callout-tip
## Key takeaways

-   The term "dispersion" is used differently in corpus linguistics and statistics
-   The difference in meaning reflects a difference in perspective
-   Corpus linguists picture the corpus as a sequence of words and understand the term as characterizing the spatial distribution of an item
-   In the statistical literature on count data modeling, the term describes the spread of a distribution of counts or occurrence rates
:::
