---
title: "Issues in random-forest modeling: Treatment of clustering variables"
description: "Corpus data often show a hierarchical structure, where observations are grouped by text/speaker or item. In this blog post, I comment on the practice of including clustering variables as predictors into a random-forest model. I demonstrate that this undermines the predictive utility of associated cluster-level predictors."
date: 2026-01-10
categories: [corpus linguistics, clustered data, bias, binary data, random forests]
citation: 
  url: https://lsoenning.github.io/posts/2026-01-10_random_forest_clustering_variable/
editor: source
---

I recently did a systematic review on the use of random-forest models in corpus research [@Soenning2026]. Perhaps unsurprisingly, 112 of the 125 RF models (90%) that entered this survey were fit to data with a grouping structure, where observations are clustered by text/speaker or item. In a considerable number of cases, the clustering variable was included into the RF model as an additional predictor variable:

- Of the 112 RF models with data clustered by speaker/text, 20 (18%) included the clustering variable into the analysis. 
- Of the 80 RF models with data clustered by item (word form or lexeme), 5 (6%) included the clustering variable as a predictor. 

In all cases, the clustering variable entered the analysis as an additional predictor, which means that the recursive partitioning scheme treated it on a par with the other variables in the model (i.e. as the equivalent of a fixed effect in a regression framework). 

In this blog post, I want to comment on the situation where a clustering variable is included in an RF model alongside an associated cluster-level predictor. This is the case if a model includes, say, (i) the clustering variable speaker as well as speaker-level predictors such as age or gender; or (ii) the clustering variable word as well as word-level predictors such as frequency or word class.

In the survey I did, around half of the RF models that included a clustering variable also included a predictor that was measured at the cluster level: 

- Of the 20 RFs including speaker as a predictor variable, 11 (55%) also had speaker/text-level predictors. Examples included: age, gender, education, generation, year of birth, ethnicity 
- Of the 5 RFs that include item as a predictor variable, 3 (60%) also feature item-level predictors. Examples included: verb ending, verb modality, frequency, gradability, semantic category

I can see why some may think that including a clustering variable into an RF model might be a way of addressing the structure of the data – however, ordinary random forests do not handle clustering variables (e.g. speaker and item) appropriately. By "ordinary", I mean RFs that do not take a mixed-effects approach to modeling clustering variables. In this blog post, I show that if these variables are included into a random-forest model, the "effect" of cluster-level predictors will be attenuated. Despite prominent precedent [@tagliamonte_baayen2012], then, this practice cannot be recommended. In this blog post, I use a simulated set of data to demonstrate this issue.

```{r}
#| code-fold: true
#| code-summary: "R setup"
#| message: false
#| warning: false

library(party)
library(permimp)
library(lattice)
library(vcdExtra)
library(tidyverse)
library(uls)
```


#### Simulate data

We start by simulating data with a binary response variable that is sensitive to speaker `age` and `gender`. The fictitious data include 50 speakers, with each individual contributing 30 observations to the data. 

```{r}
#| code-fold: true
#| code-summary: "simulate data"

# set seed for reproducibility
set.seed(11)

# number of speaker and number of observations per speaker
n_speakers <- 50
n_tokens_per_speaker <- 30

# simulate speaker intercepts (on the logit scale)
speaker_intercepts_logit <- rnorm(n_speakers, 0, 1)

# simulate speaker age and gender
speaker_age <- (runif(n_speakers) -.5) * 4
speaker_gender <- rep(c(0,1), each = n_speakers/2)

# translate speaker intercepts to the proportion scale
speaker_intercepts_prob <- plogis(
  -speaker_age*.8 + (speaker_gender - .5) + speaker_intercepts_logit)

# use speaker-specific proportion to simulate binary outcome
speaker_were <- rbinom(
  n_speakers, n_tokens_per_speaker, speaker_intercepts_prob)

# obtain empirical proportion for each speaker
speaker_were_prop <- speaker_were/n_tokens_per_speaker


# organize these data into case form (one row per observation):
d <- data.frame(
  speaker = paste0("subj_", 1:n_speakers),
  age = speaker_age,
  gender = speaker_gender,
  were = speaker_were,
  was = n_tokens_per_speaker - speaker_were
) |> 
  pivot_longer(
    cols = c("was", "were"), 
    names_to = "variant", 
    values_to = "freq")

d <- expand.dft(d, freq = "freq") 

d$variant <- factor(d$variant)
d$speaker <- factor(d$speaker)
d$gender <- factor(d$gender, labels = c("male", "female"))

# add noise variables
d$random_continuous <- rnorm(nrow(d))
d$random_binary <- rbinom(nrow(d), 1, .5)
```

@fig-data shows the patterns of association we have built into the data:

- The proportion of interest is linked to `age`: The younger the speaker, the lower the proportion.
- Speaker `gender` also plays a role, with male speakers showing a higher proportion.

```{r}
#| fig-width: 3.8
#| fig-height: 2.3
#| code-fold: true
#| code-summary: "draw graph"
#| label: fig-data
#| fig-cap: "Scatterplot showing how the proportion of interest varies with `age` and `gender`."
#| message: false
#| warning: false

p1 <- xyplot(
  speaker_were_prop ~ speaker_age, 
  groups = speaker_gender, pch = c(21,3), col = c("black", "black"), cex = c(1, .8),
  par.settings = lattice_ls, axis = axis_L, ylim = c(0,1),
  xlab = "Age", ylab = "Proportion of interest",
  scales = list(
    x = list(at = -1.5:1.5, label = rev(c(20, 40, 60, 80)), cex = .9),
    y = list(at = c(0, .5, 1), label = c("0", ".5", "1")), cex = .9),
  panel = function(x,y){
    #panel.abline(h = 0, col = "black")
    panel.points(x[speaker_gender == 0], y[speaker_gender == 0], lwd = .75, pch = 3, cex = .9, col = "black")
    panel.points(x[speaker_gender == 1], y[speaker_gender == 1], lwd = .75, pch = 21, cex = 1.1)
    
    panel.dotdiagram(speaker_were_prop[speaker_gender == 1], 
                     x_anchor = 2.6, seq_min = 0, seq_max = 1, 
                     n_bins = n_tokens_per_speaker, scale_y = .1, vertical = TRUE,
                     set_pch = 21, set_lwd = .75, set_cex = 1.1)
    
    panel.dotdiagram(speaker_were_prop[speaker_gender == 0], 
                     x_anchor = 3.6, seq_min = 0, seq_max = 1, 
                     n_bins = n_tokens_per_speaker, scale_y = .1, vertical = TRUE,
                     set_pch = 3, set_lwd = .75, set_cex = .9, set_col = "black")
    panel.text(x = c(2.6, 3.6), y = -.075, label = c("Male", "Female"), adj = 0, cex = 1, col = c("black", "black"))
  })

print(p1, position=c(0,0,.7,1))

```

We also add to the dataset two variables that represent random noise, a binary one (`random_binary`) and a continuous one (`random_continuous`). This is for two reasons:

- First, it will increase the number of predictors, making RF modeling (which includes predictor sampling) feasible
- Second, and more importantly, the variable importance scores we obtain for these noise variables are a useful point of reference for the predictors of interest. Due to the associations we have built into the model, `age` and `gender` should exceed both noise variables in predictive utility.

This is the resulting dataset:

```{r}
str(d)
```


#### Modeling

In what follows, we will examine the consequences of including the clustering variable `speaker` as an additional variable into the model. We will therefore fit and compare two models, M1 and M2:

- **M1** does not include the clustering variable `speaker`
  - formula: `variant ~ age + gender + random_continuous + random_binary`
- **M2** includes the clustering variable `speaker`
  - formula: `variant ~ age + gender + random_continuous + random_binary + speaker`

For didactic purposes, we start by fitting a conditional inference tree to the full dataset using the function `ctree()` in the `{party}` package [@party-package]. The resulting splitting scheme is shown in @fig-cit. It includes three splits: three according to `speaker`, and one according to `gender`. Surprisingly, `age` doesn't appear at all as a splitting variable, which should make us skeptical.

```{r}
#| fig-width: 10
#| fig-height: 6
#| out-width: 90%
#| label: fig-cit
#| fig-cap: "Conditional inference tree fit to the full dataset."

d_tree <- party::ctree(
  variant ~ age + gender +
    random_continuous + random_binary +
    speaker, 
  data = d)

plot(d_tree)
```

The next step is to fit two conditional random-forest models (M1 and M2) using the function `cforest()` in the `{party}` package [@party-package].

```{r}
#| eval: false

# M1
d_rf_1 <- party::cforest(
  variant ~ age + gender + 
    random_continuous + random_binary, 
  data = d,
  control = party::cforest_unbiased(
    mtry = 2, 
    ntree = 500))

# M2
d_rf_2 <- party::cforest(
  variant ~ 
    age + gender +
    random_continuous + random_binary + 
    speaker,  
  data = d,
  control = party::cforest_unbiased(
    mtry = 2, 
    ntree = 500))

```

Next, we compute conditional variable importance scores for each model, using the `permimp` package [@permimp-package]:

```{r fig.height=1.5, fig.width=3}
#| eval: false

# M1
varimp_d_rf_1 <- permimp::permimp(
  d_rf_1,
  conditional = TRUE,
  progressBar = FALSE)$values

varimp_d_rf_1 <- as.data.frame(varimp_d_rf_1)
varimp_d_rf_1$variable <- rownames(varimp_d_rf_1)


# M2
varimp_d_rf_2 <- permimp::permimp(
  d_rf_2,
  conditional = TRUE,
  progressBar = FALSE)$values

varimp_d_rf_2 <- as.data.frame(varimp_d_rf_2)
varimp_d_rf_2$variable <- rownames(varimp_d_rf_2)

```

Finally, we obtain partial dependence scores for the predictors `age`and `gender` for each model, using the `partial()` function in the `pdp` package [@pdp-package]:

```{r}
#| eval: false

# M1
pdp_gender_1 <- pdp::partial(
  d_rf_1, 
  pred.var = "gender",
  type = "classification",
  which.class = "variant.were",
  prob = TRUE)

pdp_age_1 <- pdp::partial(
  d_rf_1, 
  pred.var = "age", 
  pred.grid = data.frame(
    age = seq(-2, 2, .25)),
  type = "classification",
  which.class = "variant.were",
  prob = TRUE)


# M2
pdp_gender_2 <- pdp::partial(
  d_rf_2, 
  pred.var = "gender",
  type = "classification",
  which.class = "variant.were",
  prob = TRUE)

pdp_age_2 <- pdp::partial(
  d_rf_2, 
  pred.var = "age", 
  pred.grid = data.frame(
    age = seq(-2, 2, .25)),
  grid.resolution = 10,  
  type = "classification",
  which.class = "variant.were",
  prob = TRUE)
```


```{r}
#| eval: false
#| echo: false

saveRDS(d_rf_1, "./models/d_rf_1.rds")
saveRDS(d_rf_2, "./models/d_rf_2.rds")

saveRDS(varimp_d_rf_1, "./models/varimp_d_rf_1.rds")
saveRDS(varimp_d_rf_2, "./models/varimp_d_rf_2.rds")

saveRDS(pdp_age_1, "./models/pdp_age_1.rds")
saveRDS(pdp_age_2, "./models/pdp_age_2.rds")

saveRDS(pdp_gender_1, "./models/pdp_gender_1.rds")
saveRDS(pdp_gender_2, "./models/pdp_gender_2.rds")

```

```{r}
#| echo: false

varimp_d_rf_1 <- readRDS("./models/varimp_d_rf_1.rds")
varimp_d_rf_2 <- readRDS("./models/varimp_d_rf_2.rds")

varimp_d_rf_1$varimp_scaled <- 
  varimp_d_rf_1[,1] / max(varimp_d_rf_1[,1])

varimp_d_rf_2$varimp_scaled <- 
  varimp_d_rf_2[,1] / max(varimp_d_rf_2[,1])


pdp_age_1 <- readRDS("./models/pdp_age_1.rds")
pdp_age_2 <- readRDS("./models/pdp_age_2.rds")

pdp_gender_1 <- readRDS("./models/pdp_gender_1.rds")
pdp_gender_2 <- readRDS("./models/pdp_gender_2.rds")
```


#### Variable importance scores

We start by inspecting the variable importance scores, which are shown in the @fig-vim. For comparison, I have scaled these in such a way that each plot extends to the maximum importance score in the respective model. This rescaling does not affect our interpretation here since the relevant benchmarks are the conditional importance scores for the predictors `random_binary` and `random_continuous`. 

- For M1, the importance measures suggest that both `age` and `gender` have predictive utility, in line with the structure we have built into the data.
- For M2, it is `age` and `speaker` that emerge as predictively important; `gender`, however, ranks with the noise variables.

```{r}
#| fig-width: 3
#| fig-height: 2.3
#| code-fold: true
#| code-summary: "draw graph"
#| label: fig-vim
#| fig-cap: "Conditional variable importance for the predictors in models M1 and M2."
#| message: false
#| warning: false

xyplot(
  1~1, type = "n", xlim = c(-.2, 1.1), ylim = c(0, 12.5),
  par.settings = lattice_ls,
  xlab = "Conditional importance (scaled)",
  ylab = NULL,
  scales = list(
    x = list(at = c(0, 1), label = c("0", "Max."), cex = .9),
    y = list(
      at = c(1:5, 8:11), label = c(
        "(Continuous)", "(Binary)", "Gender", "Speaker", "Age",
        "(Continuous)", "(Binary)", "Gender", "Age"
      )), cex = .9),
  panel = function(x,y){
    panel.segments(x0 = -.2, x1 = 1.1, y0 = c(1:5, 8:11), y1 = c(1:5, 8:11), 
                   col = "grey95")
    panel.segments(x0 = 0, x1 = 0, y0 = .5, y1 = 5.5)
    panel.segments(x0 = 0, x1 = 0, y0 = 7.5, y1 = 11.5)
    panel.segments(x0 = 0, x1 = sort(varimp_d_rf_2$varimp_scaled),
                   y0 = 1:5, y1 = 1:5)
    panel.segments(x0 = 0, x1 = sort(varimp_d_rf_1$varimp_scaled),
                   y0 = 8:11, y1 = 8:11)
    panel.points(x = sort(varimp_d_rf_2$varimp_scaled), y = 1:5,
                 pch = 21, fill = "white", cex = 1)
    
    panel.points(x = sort(varimp_d_rf_1$varimp_scaled), y = 8:11,
                 pch = 19, cex = 1)
    panel.segments(x0 = 0, x1 = 1, y0 = 0, y1 = 0)
    panel.segments(x0 = 0:1, x1 = 0:1, y0 = 0, y1 = -.3)
    panel.text(x = .5, y = c(6, 12)+.2, label = c("M2", "M1"), cex = 1)
  })
```

Apparently, the inclusion of the variable `speaker` has affected the predictive utility of the variable `gender`. 


#### Partial dependence plots

To shed further light on this, we inspect partial dependence plots for the predictors `age` and `gender` based on both models. These are compared in @fig-pdp. It is clear from this figure that the inclusion of the clustering variable `speaker` has undermined the predictive power of both speaker-level variables: The partial dependence scores based on M1, which appear as open circles, are more extreme, indicating greater predictive discrimination for young vs. old and male vs. female individuals In contrast, partial dependence scores based on M2, which includes `speaker`, are flatter. 

```{r}
#| fig-width: 3.5
#| fig-height: 2.4
#| classes: preview-image
#| code-fold: true
#| code-summary: "draw graph"
#| label: fig-pdp
#| fig-cap: "Partial dependence plots for `age` and `gender` in models M1 and M2."
#| message: false
#| warning: false

p1 <- xyplot(
  1~1, type = "n", xlim = c(-2.2, 2.2), ylim = c(0,1),
  par.settings = lattice_ls, axis = axis_L,
  xlab = "", ylab = "Proportion of interest",
  scales = list(
    x = list(at = -1.5:1.5, label = rev(c(20, 40, 60, 80)), cex = .9),
    y = list(at = c(0, .5, 1), label = c("0", ".5", "1")), cex = .9),
  panel = function(x,y){
    panel.points(x = pdp_age_2$age,
                 y = pdp_age_2$yhat, type = "l")
    
    panel.points(x = pdp_age_1$age,
                 y = pdp_age_1$yhat, type = "l")
    
    panel.points(x = pdp_age_2$age,
                 y = pdp_age_2$yhat, pch = 19, cex = 1.1)
    
    panel.points(x = pdp_age_1$age,
                 y = pdp_age_1$yhat, pch = 21, fill = "white", cex = 1.1)
    panel.text(x = 1.5, y = 1-c(.55, .9), 
               label = c("M2", "M1"), cex = .9)
    panel.text(x = -2.1, y = 1.07, label = " Age", adj = 0)
    panel.text(x = c(-2.1, -9.45), y = 1.15, label = c(
      " Partial dependence plots\n", "Conditional variable\nimportance scores (rescaled)"), adj = 0, lineheight = .85)
  })

p2 <- xyplot(
  1~1, type = "n", xlim = c(2.2, .8), ylim = c(0,1),
  par.settings = lattice_ls, axis = axis_L,
  xlab = "", ylab = " ",
  scales = list(
    x = list(at = 1:2, label = c("Male", "Female"), cex = .9),
    y = list(at = c(0, .5, 1), label = c("0", ".5", "1")), cex = .9),
  panel = function(x,y){
    panel.points(x = pdp_gender_2$gender,
                 y = pdp_gender_2$yhat, type = "l")
    
    panel.points(x = pdp_gender_1$gender,
                 y = pdp_gender_1$yhat, type = "l")
    
    panel.points(x = pdp_gender_2$gender,
                 y = pdp_gender_2$yhat, pch = 19, cex = 1.1)
    
    panel.points(x = pdp_gender_1$gender,
                 y = pdp_gender_1$yhat, pch = 21, fill = "white", cex = 1.1)
    panel.text(x = 1.9, y = 1-c(.33, .63), 
               label = c("M1", "M2"), cex = .9)
    panel.text(x = 2.2, y = 1.07, label = " Gender", adj = 0)
  })

print(p1, position = c(0,-.1,.65,.86), more = TRUE)
print(p2, position = c(.6,-.1, 1,.86))

```

We can compare the "steepness" of the profiles numerically, by looking at the range of the partial dependence scores for the two models:

- `age` 
  - M1: .52 (Min: .20; Max: .72)
  - M2: .25 (Min: .32; Max: .57)
- `gender` 
  - M1: .22 (Min: .35; Max: .56)
  - M2: .10 (Min: .39; Max: .49)

```{r}
#| echo: false
#| eval: false

# age
round(range(pdp_age_2$yhat), 2)
round(range(pdp_age_1$yhat), 2)
round(diff(range(pdp_age_2$yhat)), 2)
round(diff(range(pdp_age_1$yhat)), 2)

# gender
round(range(pdp_gender_2$yhat), 2)
round(range(pdp_gender_1$yhat), 2)
round(diff(range(pdp_gender_2$yhat)), 2)
round(diff(range(pdp_gender_1$yhat)), 2)
```

#### Understanding why

To understand why the inclusion of clustering variables undermines the predictive utility of the associated cluster-level predictors, it is instructive to inspect the first split in the conditional inference tree shown in @fig-cit. The first split is made on `speaker`, and 18 individuals are assigned to the left branch, 32 to the right branch. The sample is therefore divided into two groups. The scatterplot in @fig-explain shows that this split is made at a proportion of around .55. The subset of speakers in the left branch is colored in gray, the ones in the right branch in blue.

For this split, the prediction scheme assigns predictive value to the variable `speaker` only. This is despite the fact that the two subgroups differ systematically with regard to the distribution of `age` and `gender`. This is clarified in the plots added at the top of @fig-explain:

- The bar chart shows that the right branch (blue) includes a greater proportion of male speakers (63% vs. 28% in the right branch). Recall that male speakers were simulated to have lower proportions (see @fig-data).
- The boxplot shows that the right branch (blue) also includes speakers who are, on average, younger (median age 33 vs. 62 years). Recall that younger speakers were simulated to have lower proportions (see @fig-data). 

```{r}
#| fig-width: 4
#| fig-height: 3.6
#| code-fold: true
#| code-summary: "draw graph"
#| label: fig-explain
#| fig-cap: "Explanation of the decrease in predictive utility of cluster-level variables in M2."

# speaker summary
d_summary <- d |> group_by(speaker) |> 
  dplyr::summarize(
    age = unique(age),
    gender = unique(gender),
    prop_were = mean(variant == "were")
  )

d_tree <- ctree(
  variant ~ age + gender + speaker +
    random_continuous + random_binary, 
  data = d)

speaker_split <- d_tree@tree$psplit$splitpoint

speakers_to_the_left <- levels(speaker_split)[as.logical(speaker_split)]
speakers_to_the_right <- levels(speaker_split)[as.logical(speaker_split*-1+1)]

d_summary$branch <- factor(
  c("right", "left")[as.numeric(d_summary$speaker %in% speakers_to_the_left)+1])


# draw plot
p1 <- xyplot(
  prop_were ~ age, 
  col = col_ls[c(1,3)][as.numeric(d_summary$branch)], 
  pch = c(3,1)[as.numeric(d_summary$gender)],
  cex = 1,
  data = d_summary,
  xlab = "Age", ylab = "Proportion of interest", ylim = c(0,1),
  par.settings = lattice_ls, axis = axis_L,
scales = list(
    x = list(at = -1.5:1.5, label = rev(c(20, 40, 60, 80))),
    y = list(at = c(0, .5, 1), label = c("0", ".5", "1"))),
panel = function(x,y,...){
  panel.xyplot(x,y,...)
  panel.segments(x0 = 2.5, x1 = 2.5, col = col_ls[1],
                 y0 = min(y[d_summary$branch == "left"]),
                 y1 = max(y[d_summary$branch == "left"]))
  panel.segments(x0 = 2.4, x1 = 2.5, col = col_ls[1],
                 y0 = range(y[d_summary$branch == "left"]),
                 y1 = range(y[d_summary$branch == "left"]))
  
  panel.segments(x0 = 2.5, x1 = 2.5,col = col_ls[3],
                 y0 = min(y[d_summary$branch == "right"]),
                 y1 = max(y[d_summary$branch == "right"]))
  panel.segments(x0 = 2.4, x1 = 2.5,col = col_ls[3],
                 y0 = range(y[d_summary$branch == "right"]),
                 y1 = range(y[d_summary$branch == "right"]))
  
  panel.text(x = 2.7, y = c(.25, .75), label = c(
    "Assigned to\nright branch\n(n = 32)", "Assigned to\nleft branch\n(n = 18)"
  ), lineheight = .85, adj = 0, cex = .9, col = col_ls[c(3,1)])
    })

p2 <- bwplot(reorder(branch, -age) ~ age, data = d_summary,
  par.settings = lattice_ls, box.ratio = 4,
  scales = list(x = list(draw = FALSE),
                y = list(cex = .9)),
  fill = fill_ls[c(3,1)],
  xlab = NULL,
  panel = function(x,y,...){
    panel.bwplot(x,y,...)   
    panel.text(x = 2.7, y = 3, label = "Median age", cex = .9, adj = 0)
    panel.text(x = 2.7, y = 2, label = "  Left branch: 33 years", cex = .9, adj = 0, col = col_ls[1])
    panel.text(x = 2.7, y = 1, label = "  Right branch: 62 years", cex = .9, adj = 0, col = col_ls[3])
  })

d_barchart <- data.frame(
  prop.table(
    table(d_summary$gender, d_summary$branch), margin = 2))


p3 <- barchart(
  reorder(Var2, -as.numeric(Var2)) ~ Freq, groups = Var1:Var2,
               data = d_barchart, stack = TRUE,
  par.settings = lattice_ls, box.ratio = 4,
  xlab = NULL,
  scales = list(x = list(draw = FALSE),
                y = list(cex = .9)),
  col = c(
    fill_ls[1],
    fill_ls[3], 
    "white",
    "white"),
  panel = function(x,y,...){
    panel.barchart(x,y,...)
    panel.text(x = c(.1, .88), y = 2.95, label = c("male", "female"), cex = .9)
    panel.text(x = 1.19, y = 3, label = "Proportion male", cex = .9, adj = 0)
    panel.text(x = 1.19, y = 2, label = "  Left branch: 28%", cex = .9, adj = 0, col = col_ls[1])
    panel.text(x = 1.19, y = 1, label = "  Right branch: 63%", cex = .9, adj = 0, col = col_ls[3])
  })

print(p1, position = c(0,0,.62,.6), more = TRUE)
print(p2, position = c(-.02,.625,.625,.8), more = TRUE)
print(p3, position = c(-.02,.8,.625,.975))
```

Since we have built the data, we know that part of the difference between the two subgroups of speakers (grey and blue) is systematic, i.e. attributable to `age` and `gender`. For this particular split, however, the tree records no information about systematic differences between grey and blue individuals. Instead, predictive credit is given exclusively to `speaker`. Since the right branch (blue) mostly includes younger speakers, predictive power has been robbed from `age` and instead assigned to `speaker`. And the same is true for `gender`: Since the right branch (blue) mostly includes male speakers, predictive power has also been robbed from `gender` and instead assigned to `speaker`.


#### Conclusion

Corpus data are often structured hierarchically, which poses a problem for random-forest modeling. We know from the regression context that ignoring clustering variables during data analysis is likely to yield overoptimistic models. Including clustering variables into an ordinary random-forest model has its own problems, however, one of which was discussed in the current blog post. If the data are to be analyzed with a random-forest model, the only way out of this dilemma is offered by mixed-effects random forests. For categorical response variables (i.e. classification tasks), such methods have emerged only recently [@pellagatti2021]. If you come across an implementation of these models in R (or Python), spread the word! 
